---
title: Analysis of U.S. Storm Event Data and it's Impact on Public Health and the
  Economy
author: "Gaurab Kundu"
date: "2022-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis
 In this project, we analyze the storm database taken from the U.S. National Oceanic and Atmospheric Administration (NOAA). We estimate the fatalities, injuries, property damage, and crop damage for each type of event (e.g., Flood, Typhoon, Tornado, Hail, Hurricane, etc.). Our goal is to determine which event is most harmful to US population (health) and which event has the largest economic consequences. Our analysis on Fatalities and Injuries conclude that Tornado is the most harmful event in respect to the US health (population). On the other hand, based on the Property and Cost damage, we conclude that Flood has the greatest economic consequences to the US.

# Introduction
 Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

 This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data
 The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the following link:

Storm Data [47Mb]
There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation
National Climatic Data Center Storm Events FAQ
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

# Questions
## The data analysis address the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

# Data Processing

## Loading required libraries

```{r}
library(data.table)
library(ggplot2)
```

## Loading the Data into R

```{r}
data <- read.csv("repdata_data_StormData.csv.bz2", header = TRUE, sep=",")
```

## Inspecting the Data

```{r}
colnames(data) #checking column names
```

## Subsetting the Data

We are only interested in the column related to health and economic impacts. Therefore, only the following columns are needed and we can remove the rest.

### EVTYPE
### FATALITIES
### INJURIES
### PROPDMG
### PROPDMGEXP
### CROPDMG
### CROPDMGEXP

```{r}
selection <- c('EVTYPE', 'FATALITIES', 'INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')
data <- data[, selection]
summary(data)
```

We also only need to use the data where fatalities, injuries, or damages occured.

```{r}
data <- as.data.table(data)
data <- data[(EVTYPE != "?" & (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), 
             c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```

## Converting the exponent columns (PROPDMGEXP and CROPDMGEXP)

We need to convert the exponent values from K, M, B to 1000, 1000000, and 1000000000.

```{r}
cols <- c("PROPDMGEXP", "CROPDMGEXP")
data[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

PROPDMGKey <-  c("\"\"" = 10^0, 
                 "-" = 10^0, "+" = 10^0, "0" = 10^0, "1" = 10^1, "2" = 10^2, "3" = 10^3,
                 "4" = 10^4, "5" = 10^5, "6" = 10^6, "7" = 10^7, "8" = 10^8, "9" = 10^9, 
                 "H" = 10^2, "K" = 10^3, "M" = 10^6, "B" = 10^9)
CROPDMGKey <-  c("\"\"" = 10^0, "?" = 10^0, "0" = 10^0, "K" = 10^3, "M" = 10^6, "B" = 10^9)

data[, PROPDMGEXP := PROPDMGKey[as.character(data[,PROPDMGEXP])]]
data[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

data[, CROPDMGEXP := CROPDMGKey[as.character(data[,CROPDMGEXP])] ]
data[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```

## Creating two new columns of Property Cost and Crop Cost

Combining the coefficient (mantissa) and exponent part of Property and Crop Damage.

```{r}
data <- data[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, PROPCOST = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, CROPCOST = CROPDMG * CROPDMGEXP)]
```

# Analysis

## Estimating the total of Fatalities and Injuries (Health Impacts)

In order to know the health impact, we estimate the total of Fatalities and Injuries for each event.

```{r}
Health_Impact <- data[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), TOTAL_HEALTH_IMPACTS = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]

Health_Impact <- Health_Impact[order(-TOTAL_HEALTH_IMPACTS), ]

Health_Impact <- Health_Impact[1:10, ]

head(Health_Impact, 10)
```

## Estimating the total of Property Cost and Crop Cost (Economic Impacts)

In order to know the economic impact, we estimate the total of Property Cost and Crop Cost for each event.

```{r}
Eco_Impact <- data[, .(PROPCOST = sum(PROPCOST), CROPCOST = sum(CROPCOST), TOTAL_ECO_IMPACTS = sum(PROPCOST) + sum(CROPCOST)), by = .(EVTYPE)]

Eco_Impact <- Eco_Impact[order(-TOTAL_ECO_IMPACTS), ]

Eco_Impact <- Eco_Impact[1:10, ]

head(Eco_Impact, 10)
```

# Results

## Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## Answer

We generate histogram to find the top 10 weather events that are most harmful to US population.

```{r}
Health_Consequences <- melt(Health_Impact, id.vars = "EVTYPE", variable.name = "Fatalities_or_Injuries")

ggplot(Health_Consequences, aes(x = reorder(EVTYPE, -value), y = value)) + 
  geom_bar(stat = "identity", aes(fill = Fatalities_or_Injuries), position = "dodge") + 
  ylab("Total Injuries/Fatalities") + 
  xlab("Event Type") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Top 10 US Weather Events that are Most Harmful to Population") + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Question 2: Events that have the greatest economic consequences

## Answer

We generate histogram to find the top 10 weather events that have largest cost to US economy.

```{r}
Eco_Consequences <- melt(Eco_Impact, id.vars = "EVTYPE", variable.name = "Damage_Type")

ggplot(Eco_Consequences, aes(x = reorder(EVTYPE, -value), y = value/1e9)) + 
  geom_bar(stat = "identity", aes(fill = Damage_Type), position = "dodge") + 
  ylab("Cost/Damage (in billion USD)") + 
  xlab("Event Type") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Top 10 US Weather Events that have the Greatest Economic consequences") + 
  theme(plot.title = element_text(hjust = 0.5))
```

